---
title: "Generating and validating synthetic clinical data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating and validating synthetic clinical data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Motivation

Sharing individual-level clinical data across institutions is often
prohibited by GDPR, HIPAA, and institutional review boards. Synthetic
data preserves the statistical properties of real data while
eliminating re-identification risk, enabling multi-site collaboration
without data transfer.

## Example: synthesizing patient records

```{r setup}
library(syntheticdata)
```

```{r real-data}
set.seed(42)
real <- data.frame(
  age     = rnorm(500, mean = 65, sd = 12),
  sbp     = rnorm(500, mean = 135, sd = 22),
  sex     = sample(c("Male", "Female"), 500, replace = TRUE),
  smoking = sample(c("Never", "Former", "Current"), 500,
                   replace = TRUE, prob = c(0.4, 0.35, 0.25)),
  outcome = rbinom(500, 1, 0.28)
)
head(real)
```

## Parametric synthesis (Gaussian copula)

The default method estimates marginal distributions empirically and
captures the joint dependence structure via a Gaussian copula on
normal scores. This preserves both marginal shapes and pairwise
correlations.

```{r synthesize}
syn <- synthesize(real, method = "parametric", n = 500, seed = 1)
syn
```

## Validation

`validate_synthetic()` computes four classes of metrics:

```{r validate}
val <- validate_synthetic(syn)
val
```

- **KS statistic**: distributional similarity (lower = better)
- **Correlation difference**: preservation of variable associations
- **Discriminative AUC**: can a classifier distinguish real from
  synthetic? Values near 0.5 mean indistinguishable.
- **NN distance ratio**: privacy metric. Values > 1 indicate
  synthetic records are not memorizing real individuals.

## Comparing methods

```{r compare}
syn_boot  <- synthesize(real, method = "bootstrap", seed = 2)
syn_noise <- synthesize(real, method = "noise", noise_level = 0.2, seed = 3)

val_boot  <- validate_synthetic(syn_boot)
val_noise <- validate_synthetic(syn_noise)

comparison <- rbind(
  transform(as.data.frame(val),       method = "parametric"),
  transform(as.data.frame(val_boot),  method = "bootstrap"),
  transform(as.data.frame(val_noise), method = "noise")
)
comparison[, c("method", "metric", "value", "interpretation")]
```

## Privacy-utility trade-off

Higher `noise_level` improves privacy but reduces utility:

```{r tradeoff}
results <- list()
for (nl in c(0.05, 0.1, 0.2, 0.5)) {
  s <- synthesize(real, method = "noise", noise_level = nl, seed = 1)
  v <- validate_synthetic(s)
  results <- c(results, list(data.frame(
    noise_level = nl,
    ks = v$value[v$metric == "ks_statistic_mean"],
    privacy = v$value[v$metric == "nn_distance_ratio"]
  )))
}
do.call(rbind, results)
```
